---
title: "EDA: Jena micro climate"
author: "Daniel Loos"
format: html
editor: visual
---

## Objective

What do wqe want to present at the Naturerlebniszentrum as an interactive chart?

## Load data

```{r}
library(tidyverse)
library(lubridate)
library(ggpubr)
library(leaflet)

theme_set(theme_classic())

sites <- read_csv("data/sites.csv")
load("data/15_data.all.2022.RData")
observations <- data.all

observations_daily <- 
  data.all |>
  mutate(date.time = date.time |> date() |> as_datetime()) |>
  group_by(WZE_ID, date.time) |>
  summarise(across(starts_with("T"), partial(mean, na.rm = TRUE))) |>
  left_join(sites)
```

## Seasonal impacts

```{r}
observations_daily |>
  left_join(sites) |>
  filter(! is.na(T.air_150_cm)) |>
  ggplot(aes(date.time, elevation)) +
    geom_point(aes(color = T.air_150_cm), size = 3) +
    scale_color_viridis_c(option = "A") +
    labs(
      x = "Zeit", y = "Höhe über NN [m]",
      color = "Lufttemperatur im Tagesmittel\nin 1.5m Höhe [°C]"
    )
```

-   no clear pattern

-   little differences in height

-   sites are grouped in 3 elevation groups

```{r}
observations_daily |>
  ggplot(aes(T.air_150_cm, elevation)) +
    geom_point() +
    stat_cor()
```

no direct correlation between air temperature and elevation

```{r}
leaflet(data = sites) |>
  addTiles() |>
  addMarkers(
    ~Longitude_WGS84_E, ~Latitude_WGS84_N,
    label = ~ str_glue("{WZE_ID}, {elevation}")
    )
```

Lets take 2 close sites at different elevations:

```{r}
observations_daily |>
  filter(WZE_ID %in% c("KM1", "K2")) |>
  ggplot(aes(date.time, T.air_150_cm, color = WZE_ID)) +
    geom_line()
```

```{r}
observations_daily |>
  ggplot(aes(date.time, T.air_150_cm, color = elevation)) +
    geom_line() +
    scale_color_viridis_c()
```

-   no big seasonal change

## Within day changes

```{r}
observations <- data.all
```

```{r}
observations |>
  filter(as_date(date.time) == as.Date("2022-10-01")) |>
  ggplot(aes(date.time, T.air_150_cm, color = elevation, group = WZE_ID)) +
    geom_line() +
    scale_color_viridis_c()
```

-   there is sub daily signal: lower sites have more extreme temperatures

```{r}
temp_stats <-
  observations |>
  mutate(date = floor_date(date.time, "day")) |>
  group_by(WZE_ID, date) |>
  summarise(
    min_temp = min(T.air_150_cm),
    max_temp = max(T.air_150_cm),
    temp_span = max_temp - min_temp
  ) |>
  left_join(sites) |>
  # mutate(elevation_group = cut(elevation, seq(min(elevation), max(elevation), 5)))
  mutate(elevation_group = case_when(
    elevation > 320 ~ "Berg",
    TRUE ~ "Tal"
  ))

temp_stats_groups <-
  temp_stats |>
  group_by(date, elevation_group) |>
  summarise(temp_span = mean(temp_span, na.rm = TRUE))

temp_stats |>
  filter(WZE_ID == "EH2") |>
  ggplot(aes(date, temp_span, fill = elevation_group)) +
    stat_smooth(data = temp_stats_groups, method = "loess", color = NA) +
    geom_point(size = 0.5) +
    stat_smooth(method = "loess", color = "black", se = FALSE) +
    scale_color_manual(values = c(Berg = "#283618", Tal = "#b5838d")) +
    scale_fill_manual(values = c(Berg = "#283618", Tal = "#b5838d")) +
    guides(col = "none") +
    labs(
      title = "Extreme Temperaturen im Tal",
      x = "Zeit [Tag]",
      y = "Temperaturspanne im Tagesverlauf [°C]",
      fill = "Höhe"
    )
```
